Downstream analyses of whole-genome bisulfite-sequencing data

### 2014-12-04 Rmarkdown/wgbs_downstream_analyses.Rmd
Added methylation entropy

- `QDMR` \citep{Zhang:2011dp} and `CpG_MPs` \citep{Su:2012hl}, developed by the same research group, are methods for identifying DMRs using a form of methylation entropy from MeDIP-chip and BS-seq, respectively.
- \citet{Xie:2011cy} introduced the concept of methylation entropy for BS-seq.

While mathematically based on the idea of Shannon entropy \citep{Shannon:IH1_srIe}, the interpretation of methylation entropy has much in common with the physical concept of "disordered/heterogeneity"

@@

@@

A common thread in Section \ref{sec:downstream_analyses_of_m-tuples} is that while interesting methods have been proposed to address these questions, the software implementations are somewhat lacking. This in part motivated me to begin developing `MethylationTuples`, an R package for managing, analysing and visualising methylation patterns at m-tuples.

- I'm using to address co-methylation
- Could be used for entropy, epipolymorphism and epialleles.
- Could implement Fang or Peng's method for ASM

>>>>

There are several definitions and uses of _methylation entropy_ in the literature. \citet{Xie:2011cy} introduced the concept for bisulfite-sequencing data as a way of quantifying the variability of methylation patterns at an m-tuple. For a given 4-tuple, $(i, \ldots, i + 3)$, \citet{Xie:2011cy} analyse the set of reads aligned to it, $\{z: z \in \mathcal{R}_{(i, \ldots, i + 3} \}$, to compute the methylatione entropy. On the one hand, if we only observe a single unique methylation pattern then the m-tuple has the minimum methylation entropy of zero. On the other hand, if we observe all possible $2^{4} = 16$ methylation patterns then the m-tuple has the maximum methylation entropy of 1. Depending on the frequency of the patterns, we obtain intermediate methylation values ($0 < \text{methylation entropy} < 1)$. Of course, methylation entropy need not be limited to 4-tuples. The choice of the size of the tuples depends on the regions of interest, the read lengths and the sequencing coverage.

The natural interpretation of methylation entropy as a measure of 'disorder' allows us to quantify how heterogeneous DNA methylation is at a locus \citep{Xie:2011cy, He:2013cj}[^other_definitions]. These methylation entropies can be analysed to identify heterogeneous regions of the genome or perhaps tested for an association with a phenotype.

[^other_definitions]: This is closely related to the idea of identifying epialleles, for which methylation entropy has also played a role \citep{Li:2014ei} and which I discuss in Section \ref{sec:epialleles}. Methylation entropy has also been used to identify differential methylation, however, I do not discuss this further since this uses a different definition that is not based on analysing methylation patterns at m-tuples \citep{Zhang:2011dp, Su:2012hl}.

While \citet{Xie:2011cy} do not provide software implementing the method, a subsequent paper from some of the same researchers introduces the `DMEAS` software to do so \cite{He:2013cj} . Unfortunately, `DMEAS` is only available as a Windows binary or as a Perl script that itself is only available as PDF file, greatly limiting its usefulness.

While I have not yet done so, it is not difficult to add the calculation of methylation entropies to the `MethylationTuples` software. The input for such an analysis of methylation entropy is a list of m-tuples and the associated frequencies of each methylation pattern, exactly what `methtuple` produces. The formula for computing methylation entropies is a simple sum of ratios and log-ratios, easily implemented in R, which `MethylationTuples` is written in.

@@

- methclone
- Conrad (MPFE)
- How it differs to Emma Whitelaw's definition

@@

A common thread in Section \ref{sec:downstream_analyses_of_m-tuples} is that while interesting methods have been proposed to address these questions, the software implementations are rather lacking. `MethylationTuples` is an R package for managing, analysing and visualising methylation patterns at m-tuples.  While initially developed to support my research into co-methylation (Chapter \ref{chap:co-methylation}), `MethylationTuples`, in conjunction with `methtuple`, can also be used to do many of the above analyses.

- I'm using it to address co-methylation
- Could be used for entropy, epipolymorphism and epialleles.
- Could implement Fang or Peng's method for ASM, but less readily the "heterozygous SNP" method
